---
title: "Prioritizing-potential-aquaculture"
author: "Ixel"
date: "11/31/2025"
format: 
  pdf:
    fig-pos: "H"
    geometry: margin= 0.5in
    code-fold: false
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
editor: visual
---

What Exclusive Economic Zones (eez) on the west Coast of the US are best suited to developing marine aquaculture for several species of oysters and California datemussel.

# Set-Up

```{r, message=FALSE}
#| output: false

# Load in Libraries
library(here)
library(testthat)
library(tidyverse)
library(sf)
library(kableExtra)
library(sp)
library(stars) 
library(tmap)
library(terra)
library(viridisLite)
```

```{r,message = FALSE}
#| output: false
# Prepare Data

# Bathymetry
depth <- read_stars(here("data/depth.tif"))

# Exclusive Economic Zones (eez)
eez <- st_read(here("data/wc_regions_clean.shp"))

# Sea Surface Temperature (SST) load as a list of .tiff files 
sst <- list.files(path= "data",
                  pattern= "average_annual", # file name matching pattern
                   full.names = TRUE) # Reference entire file name matching this pattern

# Stack all rasters files in folder using rast()
sst <- rast(sst)
names(sst) #checking list
```

Check that the datasets have the same coordinate reference system

```{r, message=FALSE}

# Check and align CRS for multiple datasets
if (st_crs(depth) == st_crs(eez)) {
  print("Coordinate reference system match!")
  
} else{
  warning("CRS mismatch between depth and eez")
  # tranfgotm data to match
  depth <- st_transform(depth, st_crs(eez))

  }
   

if (st_crs(eez)$wkt != crs(sst)) {
  warning("CRS mismatch between eez and sst")
  sst <- project(sst, st_crs(eez)$wkt)  
}
```



```{r}
  
# Calculate the SST average 
mean_sst <- app(sst[[c("average_annual_sst_2008",
                         "average_annual_sst_2009", 
                         "average_annual_sst_2010",
                         "average_annual_sst_2011", 
                         "average_annual_sst_2012")]], 
                  fun = mean, na.rm = TRUE)

# Convert average SST from Kelvin to Celsius
sst_mean_celsius <- mean_sst - 273.15
  
```

# Process Data

To prepare the data for suitability analysis, SST and depth rasters 
must match in resolution, extent, and coordinate reference system. 
The five annual SST rasters (2008–2012) were averaged into a single 
mean raster using `app()` and converted from Kelvin to Celsius. The 
bathymetry raster was then cropped to the SST extent and resampled 
to match its resolution using a nearest neighbor approach.

```{r}
# First convert depth stars to SpatRaster
depth_rast <- rast(depth)
# crop depth to match SST  extent
depth_crop <- crop(depth_rast, sst_mean_celsius)

# Resample depth_crop to match SST resolution, use method near
depth_resample <- resample(depth_crop,
                           sst_mean_celsius, 
                           method = "near")
```

```{r}
#| output: true

#Check resolution match 
if (all(dim(depth_resample) == dim(sst_mean_celsius)) &&
    isTRUE(all.equal(res(depth_resample), res(sst_mean_celsius))) &&
    crs(depth_resample) == crs(sst_mean_celsius) &&
    ext(depth_resample) == ext(sst_mean_celsius)) {
  cat("✓ All raster properties match!\n")
} else {
  cat("✗ Some properties do not match\n")
}

# Match depth and SST CRS's
crs(depth_resample) <- crs(sst_mean_celsius)


cat("✓ Successfully stacked!\n")
```

# Find suitable locations for Oysters

To find suitable locations for marine aquaculture, we’ll need to find locations that are suitable in terms of both SST and depth. **Oyster** research has show that oysters need the following conditions for optimal growth: - Sea surface Reclassify SST and depth data into locations that are suitable for oysters hint: set suitable values to 1 and unsuitable values to 0 find locations that satisfy both SST and depth conditions

```{r}
# Reclassify SST (11-30°C suitable for oysters)
sst_suitable <- app(sst_mean_celsius, fun = function(x){
  ifelse(x >= 11 & x <= 30,1,0) # if temperate is between 11- 30 °C, return 1, else 0
})

# Reclassify Suitable depth range (0-70 meters for oysters)
depth_suitable <- app(depth_resample, fun = function(x){
  ifelse(x >= -70 & x <= 0,1,0) # select depths from 0 to 70 meters below surface.
})

# Suitable locations for both conditions
#Use lapp() to multiply the two suitability raster

suitable_locations <- lapp(c(sst_suitable, depth_suitable), 
                          fun = function(sst, depth) {
                             sst * depth
                           })

#suitable_locations <- sst_suitable * depth_suitable

```

# Determine the most suitable EEZ

We want to determine the total suitable area within each EEZ in order to rank zones by priority. To do so, we need to find the total area of suitable locations within each EEZ.

Select suitable cells within West Coast EEZ's find area of grid cells find the total suitable area within each EEZ hint: it might be helpful to rasterize the EEZ data

```{r}
# Rasterize the EEZ data (creates zone IDs)
eez_raster <- rasterize(eez, suitable_locations, field = "rgn")

# Mask suitable locations to EEZ boundaries

# keep only raster cells that fall within the EEZ polygons
suitable_eez <- mask(suitable_locations, eez) 

#Calculate cell area with cellSize()

# calculate actual area of each raster cell
cell_area_km2 <- cellSize(suitable_locations, unit = "km") 

# Calculate suitable area per cell
suitable_area <- suitable_eez * cell_area_km2

# Sum suitable are by EEZ zonal statistics 
eez_suitable_area <- zonal(suitable_area, eez_raster, fun = "sum", na.rm = TRUE)
colnames(eez_suitable_area) <- c("eez_id", "suitable_area_km2")
```

```{r}
# join with eez data
eez$suitable_area_km2 <- eez_suitable_area$suitable_area_km2[match
                                                             (eez$rgn, 
                                                               eez_suitable_area$eez_id)]

# Rank EEZ data
eez_ranked <- eez[order(-eez$suitable_area_km2), ]

summary_table <- data.frame(
  Rank = 1:nrow(eez_ranked),
  EEZ_Name = eez_ranked$rgn_key,
  Suitable_Area_km2 = round(eez_ranked$suitable_area_km2, 2),
  Percent_of_Total = round(100 * eez_ranked$suitable_area_km2 / 
                           sum(eez_ranked$suitable_area_km2), 1)
)

# Create a Table
eez_table <- kable(summary_table,
      caption = "EEZ Rankings by Suitable Aquaculture Area",
      col.names = c("Rank", "EEZ Region", "Suitable Area (km²)", "% of Total"),
      align = c("c", "l", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#4CAF50", color = "white") %>%
  row_spec(1, bold = TRUE, background = "#E8F5E9")  # Highlight top rank

eez_table
```

```{r, message=FALSE}
#| fig.width: 10
#| fig.height: 8

# Map with tmap basic
tm_basemap("Esri.OceanBasemap", alpha = 0.7) +
tm_shape(eez_ranked) +
  tm_polygons("suitable_area_km2",
              palette = "-mako",
               style = "cont",
              title = "Suitable Area (km²)",
              border.col = "black",
              border.lwd = 0.5) +
  tm_layout(main.title = "Suitable Oyster Aquaculuture Area by EEZ",
            main.title.size = 1.2,
            legend.outsides = TRUE,
            frame = FALSE) + 
    tm_text("rgn", # Label by region
          size = .8, # Adjust size
          col = "white", # Adjust text color
          fontface = "bold", # Labels are bolded
          xmod = -.5) + # Adjust .5 from the left
  tm_compass(type = "4star", position = c("right", "top")) +
  tm_scalebar(position = c("left", "bottom")) 
  
 
```

# Workflow for other species using a function

```{r}

find_suitable_aquaculture <- function(sst_raster, 
                                      depth_raster, 
                                      eez_sf, 
                                      sst_min, 
                                      sst_max, 
                                      depth_min, 
                                      depth_max,
                                      species_name) {
  
# Reclassify SST
sst_suitable <- app(sst_raster, fun = function(x) {
    ifelse(x >= sst_min & x <= sst_max, 1, 0)
  })
  
# Reclassify depth
depth_suitable <- app(depth_raster, fun = function(x) {
    ifelse(x >= depth_min & x <= depth_max, 1, 0)
  })
  
# Find suitable locations (both conditions)
suitable_locations <- sst_suitable * depth_suitable
names(suitable_locations) <- species_name 

# Mask to EEZ
suitable_in_eez <- mask(suitable_locations, eez_sf)
  
# Rasterize EEZ
eez_rasterize <- rasterize(eez_sf, suitable_locations, field = "rgn")
  
# Calculate cell areas
cell_area_km2 <- cellSize(suitable_locations, unit = "km")
 

# Calculate suitable area per cell (commented out for now)
suitable_area <- suitable_in_eez * cell_area_km2
  

# Sum by EEZ zone
eez_suitable_area <- zonal(
  suitable_area, 
  eez_rasterize, 
  fun = "sum", 
  na.rm = TRUE)

colnames(eez_suitable_area) <- c("EEZ_ID", "Suitable_Area_km2")
  
# Join with EEZ data and rank
eez_sf$suitable_area_km2 <-
  eez_suitable_area$Suitable_Area_km2[
    match(eez_sf$rgn, eez_suitable_area$EEZ_ID)]
  eez_ranked <- eez_sf[order(-eez_sf$suitable_area_km2), ]
  
# Create summary table
summary_table <- data.frame(
   Rank = 1:nrow(eez_ranked),
    EEZ = eez_ranked$rgn_key,
    Suitable_Area_km2 = round(eez_ranked$suitable_area_km2, 2)
  )
  
# Return results as a list
results <- list(
   species = species_name,
  sst_suitable = sst_suitable,
   depth_suitable = depth_suitable,
   suitable_locations = suitable_locations,
   suitable_in_eez = suitable_in_eez,
   eez_ranked = eez_ranked,
    summary_table = summary_table
  )
  
 return(results)
}
```

# Apply function to species of choice

## Ophiodon elongatus (Girard, 1854)

- Common Name: Lingcod
- High commercial and recreational value; key West Coast groundfish fishery
  - Depth range: 0–100 meters (most individuals; max 475 m)
  - Temperature: 7–18°C
- Source: FishBase

```{r}
# Lingcod suitable habitat
lingcod_results <- find_suitable_aquaculture(
  sst_raster = sst_mean_celsius,
  depth_raster = depth_resample,
  eez_sf = eez,
  sst_min = 7,
  sst_max = 18,
  depth_min = -100,
  depth_max = 0,
  species_name = "Lingcod"
)
# Access result as a table
kable(lingcod_results$summary_table)
```

```{r, message=FALSE}
#| out-width: 90%
#| fig-width: 8  # This determines file resolution
#| fig-height: 6 # This determines file resolution


# Plot Mussel EEZ
tm_basemap("Esri.OceanBasemap", alpha = 0.7) +
#tm_shape(mussel_results$suitable_in_eez) +
  tm_shape(lingcod_results$eez_ranked) +
  tm_polygons("suitable_area_km2",
              palette = "-mako",
              style = "cont",
              title = "Suitable Area (km²)",
              border.col = "black",
              border.lwd = 0.5) +
  tm_layout(main.title = "Suitable Lingcod Aquaculture Area by EEZ",
            main.title.size = 1.2,
            legend.outside = TRUE,
            frame = FALSE) +
  tm_text("rgn", # Label by region
          size = .7, # Adjust size
          col = "white", # Adjust text color
          fontface = "bold", # Labels are bolded
          xmod = -.5) + # Adjust .5 from the left
  tm_compass(type = "4star", 
           position = c("right", "top"),
           size = 2) +
tm_scale_bar(position = c("left", "bottom"))
```

# Reflections

This analysis identified suitable locations for oyster and Lingcod 
aquaculture along the U.S. West Coast EEZs by integrating mean sea 
surface temperature (SST, 2008–2012) and bathymetric data. Reclassifying 
continuous rasters into binary suitability layers and combining them via 
map algebra allowed environmental constraints to be evaluated jointly, 
while zonal statistics quantified suitable area within each jurisdictional 
boundary.

For oysters (11–30°C, 0–70 m), Central California (CA-C) ranked highest 
with 4,940 km² of suitable area, likely reflecting its warmer nearshore 
SSTs and broad shallow shelf. Northern California (CA-N) ranked lowest 
at 454 km², suggesting its cooler upwelling-influenced waters and 
limited shelf area reduce suitability.

Lingcod (7–18°C, 0–100 m) showed a strikingly different spatial pattern. 
Washington ranked first with 10,569 km², followed by Oregon (8,318 km²) 
and Central California (7,776 km²). This northward shift reflects 
Lingcod's cold-water preference, the cooler SSTs of Washington and 
Oregon that limit oyster suitability are well within Lingcod's optimal 
range. The contrast between species highlights how species-specific 
environmental tolerances drive fundamentally different regional 
priorities for aquaculture development.

Several limitations apply to both analyses. SST and depth alone cannot 
capture substrate type, Lingcod strongly prefer rocky reef habitat, 
which this analysis cannot resolve. Additionally, marine protected areas, 
tribal waters, shipping lanes, and harmful algal bloom frequency are not 
accounted for and would substantially refine site selection in practice. 
Future analyses could incorporate substrate classification and 
multi-criteria decision frameworks to better reflect real-world 
aquaculture constraints.

# Cite

Hall, S. J., Delaporte, A., Phillips, M. J., Beveridge, M. & O’Keefe, M. Blue Frontiers: Managing the Environmental Costs of Aquaculture (The WorldFish Center, Penang, Malaysia, 2011).

Gentry, R. R., Froehlich, H. E., Grimm, D., Kareiva, P., Parke, M., Rust, M., Gaines, S. D., & Halpern, B. S. Mapping the global potential for marine aquaculture. Nature Ecology & Evolution, 1, 1317-1324 (2017).︎

GEBCO Compilation Group (2022) GEBCO_2022 Grid (doi:10.5285/e0f0bb80-ab44-2739-e053-6c86abc0289c)

Commercially Important Species Occurring in United States (contiguous states). (2025). Sealifebase.ca. https://www.sealifebase.ca/country/CountryChecklist.php?c_code=840&vhabitat=commercial

Palomares, M.L.D. and D. Pauly. Editors. (2026). FishBase. www.fishbase.org

Oliver, R. (2025, November 25). Homework Assignment 4. Github.io. https://eds-223-geospatial.github.io/assignments/HW4.html
